Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-20 article.Rmd
Minor stylistic changes

Our analysis focuses on one of the bedrocks of the historical profession: scholarly research. We begin with the history dissertation, often the defining scholarly output of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the "training" phase of the historical profession. Data supplied by ProQuest provided roughly eighty thousand PhD dissertations completed in the United States.^[ProQuest's dataset does not permit us to limit dissertations just to dissertations produced in history departments. These dissertations include dissertations *about* history. For a justification of this approach, see a [series of blog posts](http://lincolnmullen.com/research/history-dissertations/) exploring the dataset. The American Historical Association's dataset of dissertations relies on reporting from departments and does not have the chronological range of the ProQuest data. Nevertheless, applying the same method to the AHA's dataset does not provide any reason for thinking that the proportion authorship by gender over time is substantially different when limited to just history department dissertations. American Historical Association, *Directory of History Dissertations*, https://secure.historians.org/pubs/dissertations/. Data export provided to author by Liz Townsend, March 3, 2014.] Identifying the gender of their authors gives a sense for the number of women and men who are completing PhD-level training in history each year. Our program uses the year a historian wrote his or her dissertation to estimate a period for when they might have been born: we have assumed that a historian was between 25 and 45 years old when completing the dissertation. Using this temporal information, we are better able to infer their gender and chart how the larger representation of women and men changes year-by-year.

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The "coin of the realm" for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. This process often takes the form of book reviews published in academic journals. One of the leading journals in the historical profession is the *American Historical Review*, the flagship journal of the American Historical Association. Along with a few dozen articles, the AHR publishes five issues each year that contain, in aggregate, roughly one thousand books reviews covering (in its own words) "every major field of historical study." The AHR is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, and length of the AHR makes it an ideal source to analyze the historical profession on a large scale.

>>>>

Our analysis focuses on one of the bedrocks of the historical profession: scholarly research. We begin with the history dissertation, often the defining scholarly output of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the "training" phase of the historical profession. To study dissertations about history, we used data supplied by ProQuest for roughly eighty thousand PhD dissertations completed in the United States.^[ProQuest's dataset does not permit us to limit dissertations just to dissertations produced in history departments. These dissertations include dissertations *about* history. For a justification of this approach, see a [series of blog posts](http://lincolnmullen.com/research/history-dissertations/) exploring the dataset. The American Historical Association's dataset of dissertations relies on reporting from departments and does not have the chronological range of the ProQuest data. Nevertheless, applying the same method to the AHA's dataset does not provide any reason for thinking that the proportion of authorship by gender over time is substantially different when limited to just history department dissertations. American Historical Association, *Directory of History Dissertations*, https://secure.historians.org/pubs/dissertations/. Data export provided to author by Liz Townsend, March 3, 2014.] Identifying the gender of their authors identifies the approximate number of women and men who are completing PhD-level training in history each year. Our program uses the year a historian wrote his or her dissertation to estimate a period for when they might have been born: we have assumed that a historian was between 25 and 45 years old when completing the dissertation. Using this temporal information, we are better able to infer their gender and chart how the larger representation of women and men changes year-by-year.

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The "coin of the realm" for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. This process often takes the form of book reviews published in academic journals. One of the leading journals in the historical profession is the *American Historical Review*, the flagship journal of the American Historical Association. Along with a few dozen articles published in its five issues each year, the AHR publishes roughly one thousand books reviews per year covering (in its own words) "every major field of historical study." The AHR is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, and length of the AHR makes it an ideal source to analyze the historical profession on a large scale.
